---
title: "Data visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 3.1.1 Prerequisites
```{r}
library(tidyverse)
```

```{r}
ggplot2::mpg
?mpg
```

## Creating a ggplot

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y= hwy))
```

## Exercises

### Exercise 3.2.1
```{r}
ggplot(data = mpg)
```

This code creates an empty plot. The ggplot() function creates the background of the plot, but since no layers were specified with geom function, nothing is drawn.

### Exercise 3.2.2

```{r}
nrow(mtcars)
ncol(mtcars)
glimpse(mtcars)
```

There are 32 rows and 11 columns in the mtcars data frame.

### Exercise 3.2.3

```{r}
?mpg
```

The drv variable is a categorical variable which categorizes cars into front-wheels, rear-wheels, or four-wheel drive.

### Exercise 3.2.4

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cyl))
```

### Exercise 3.2.5

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```

A scatter plot is not a useful display of these variables since both drv and class are categorical variables. Since categorical variables typically take a small number of values, there are a limited number of unique combinations of (x, y) values that can be displayed. In this data, drv takes 3 values and class takes 7 values, meaning that there are only 21 values that could be plotted on a scatterplot of drv vs. class. In this data, there 12 values of (drv, class) are observed.

# Aesthetic mappings

```{r}
# Color aesthetic
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

# Size aesthetic
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))

# Transparency aesthetic
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Shape aesthetic
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

### 3.3.1 Exercises

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The argumentcolour = "blue" is included within the mapping argument, and as such, it is treated as an aesthetic, which is a mapping between a variable and a value. In the expression, colour = "blue", "blue" is interpreted as a categorical variable which only takes a single value "blue".

### Exercise 3.3.2

```{r}
mpg
```

The following list contains the categorical variables in mpg:
model
trans
drv
fl
class

The following list contains the continuous variables in mpg:
displ
year
cyl
cty
hwy

Those with <chr> above their columns are categorical, while those with <dbl> or <int> are continuous. 

### Exercise 3.3.3

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = cty))
```

Instead of using discrete colors, the continuous variable uses a scale that varies from a light to dark blue color.
When a continuous value is mapped to shape, it gives an error. Though we could split a continuous variable into discrete categories and use a shape aesthetic, this would conceptually not make sense. A numeric variable has an order, but shapes do not. It is clear that smaller points correspond to smaller values, or once the color scale is given, which colors correspond to larger or smaller values. But it is not clear whether a square is greater or less than a circle.

### Exercise 3.3.4

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = hwy, size = displ))

```

In the above plot, hwy is mapped to both location on the y-axis and color, and displ is mapped to both location on the x-axis and size. The code works and produces a plot, even if it is a bad one. Mapping a single variable to multiple aesthetics is redundant.

### Exercise 3.3.5

```{r}
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg), shape = 21, color = "black", fill = "white", size = 5, stroke = 5)
```

Stroke changes the size of the border for shapes (21-25). These are filled shapes in which the color and size of the border can differ from that of the filled interior of the shape.

### Exercise 3.3.6

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

Aesthetics can also be mapped to expressions like displ < 5. The ggplot() function behaves as if a temporary variable was added to the data with values equal to the result of the expression. In this case, the result of displ < 5 is a logical variable which takes values of TRUE or FALSE.

# Facets

```{r}
# Facet Wrap
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

```{r}
# Facet Grid
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) +
  facet_grid(. ~ cyl)
```

### 3.5.1 Exercises

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cty)
```

The continuous variable is converted to a categorical variable, and the plot contains a facet for each distinct value.

### Exercise 3.5.2

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty)) +
  facet_grid(drv ~ cyl)
```

The empty cells (facets) in this plot are combinations of drv and cyl that have no observations. These are the same locations in the scatter plot of drv and cyl that have no plots.

### Exercise 3.5.3

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

The symbol . ignores that dimension when faceting. For example, drv ~ . facet by values of drv on the y-axis. While, . ~ cyl will facet by values of cyl on the x-axis.

### Exercise 3.5.4

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

Advantages of encoding class with facets instead of color include the ability to encode more distinct categories. For me, it is difficult to distinguish between the colors of "midsize" and "minivan".

Given human visual perception, the max number of colors to use when encoding unordered categorical (qualitative) data is nine, and in practice, often much less than that. Displaying observations from different categories on different scales makes it difficult to directly compare values of observations across categories. However, it can make it easier to compare the shape of the relationship between the x and y variables across categories.

Disadvantages of encoding the class variable with facets instead of the color aesthetic include the difficulty of comparing the values of observations between categories since the observations for each category are on different plots. Using the same x- and y-scales for all facets makes it easier to compare values of observations across categories, but it is still more difficult than if they had been displayed on the same plot. Since encoding class within color also places all points on the same plot, it visualizes the unconditional relationship between the x and y variables; with facets, the unconditional relationship is no longer visualized since the points are spread across multiple plots.

The benefits encoding a variable through facetting over color become more advantageous as either the number of points or the number of categories increase. In the former, as the number of points increases, there is likely to be more overlap.

It is difficult to handle overlapping points with color. Jittering will still work with color. But jittering will only work well if there are few points and the classes do not overlap much, otherwise, the colors of areas will no longer be distinct, and it will be hard to pick out the patterns of different categories visually. Transparency (alpha) does not work well with colors since the mixing of overlapping transparent colors will no longer represent the colors of the categories. Binning methods use already color to encode density, so color cannot be used to encode categories.

As noted before, as the number of categories increases, the difference between colors decreases, to the point that the color of categories will no longer be visually distinct.

### Exercise 3.5.5

```{r}
?facet_wrap
```


The arguments nrow (ncol) determines the number of rows (columns) to use when laying out the facets. It is necessary since facet_wrap() only facets on one variable.

The nrow and ncol arguments are unnecessary for facet_grid() since the number of unique values of the variables specified in the function determines the number of rows and columns.

### Exercise 3.5.6

There will be more space for columns if the plot is laid out horizontally (landscape).


# Geometric objects

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg)+
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv))
```

```{r}
# Display multiple geoms on same plot
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth()
```

```{r}
# Display just a subset of the mpg dataset, the subcompact cars
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

### 3.6.1 Exercises

line chart: geom_line()
boxplot: geom_boxplot()
histogram: geom_histogram()
area chart: geom_area()

### Exercise 3.6.2

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

This code produces a scatter plot with displ on the x-axis, hwy on the y-axis, and the points colored by drv. There will be a smooth line, without standard errors, fit through each drv group.

### Exercise 3.6.3

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(show.legend = FALSE)
```


In the chapter, the legend is suppressed because with three plots, adding a legend to only the last plot would make the sizes of plots different. Different sized plots would make it more difficult to see how arguments change the appearance of the plots. The purpose of those plots is to show the difference between no groups, using a group aesthetic, and using a color aesthetic, which creates implicit groups. In that example, the legend isn’t necessary since looking up the values associated with each color isn’t necessary to make that point.

### Exercise 3.6.4

It adds standard error bands to the lines.

### Exercise 3.6.5

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

No. Because both geom_point() and geom_smooth() will use the same data and mappings. They will inherit those options from the ggplot() object, so the mappings don’t need to specified again.

### Exercise 3.6.6

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(mapping = aes(group = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(mapping = aes(group = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = drv)) +
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = drv)) +
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, fill = drv)) +
  geom_point(shape = 21, color = "white", size = 5, stroke = 3)
```

# Statistical transformations

```{r}
# display a bar chart of count
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

```{r}
# display a bar chart of proportion
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

```{r}
# Stat summary
ggplot(data = diamonds) +
  stat_summary(mapping = aes(x = cut, y = depth),
           fun.ymin = min,
           fun.ymax = max,
           fun.y = median)
```

# 3.7.1 Exercises

```{r}
# Stat summary with geom_pointrange
ggplot(data = diamonds) +
  geom_pointrange(mapping = aes(x = cut, y = depth),
                  stat = "summary",
                  fun.ymin = min,
                  fun.ymax = max,
                  fun.y = median)
```

The default geom for stat_summary() is geom_pointrange(). The default stat for geom_pointrange() is identity() but we can add the argument stat = "summary" to use stat_summary() instead of stat_identity().

# Exercise 3.7.2

The geom_col() function has different default stat than geom_bar(). The default stat of geom_col() is stat_identity(), which leaves the data as is. The geom_col() function expects that the data contains x values and y values which represent the bar height.

The default stat of geom_bar() is stat_bin(). The geom_bar() function only expects an x variable. The stat, stat_bin(), preprocesses input data by counting the number of observations for each value of x. The y aesthetic uses the values of these counts.

# Exercise 3.7.3

They tend to have their names in common, stat_smooth() and geom_smooth(). However, this is not always the case, with geom_bar() and stat_count() and geom_histogram() and geom_bin() as notable counter-examples. Also, the pairs of geoms and stats that are used in concert almost always have each other as the default stat (for a geom) or geom (for a stat).

# Exercise 3.7.4

The function stat_smooth() calculates the following variables:

y: predicted value
ymin: lower value of the confidence interval
ymax: upper value of the confidence interval
se: standard error
The “Computed Variables” section of the stat_smooth() documentation contains these variables.

The parameters that control the behavior of stat_smooth() include

method: the method used to
formula: the formula are parameters such as method which determines which method is used to calculate the predictions and confidence interval, and some other arguments that are passed to that.
na.rm

# Exercise 3.7.5

If group = 1 is not included, then all the bars in the plot will have the same height, a height of 1. The function geom_bar() assumes that the groups are equal to the x values since the stat computes the counts within the group.


# Position adjustments

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, color = cut))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

```{r}
# Identity position adjustement
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "identity")

ggplot(data = diamonds, mapping = aes(x = cut, color = clarity)) +
  geom_bar(fill = NA, position = "identity")
```

```{r}
# Fill position adjustement
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

```{r}
# Dodge position adjustement
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

```{r}
# Jitter position adjustement
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

# 3.8.1 Exercises

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = "jitter")
```

There is overplotting because there are multiple observations for each combination of cty and hwy values.
The relationship between cty and hwy is clear even without jittering the points but jittering shows the locations where there are more observations.

# Exercise 3.8.2

From the geom_jitter() documentation, there are two arguments to jitter:

width controls the amount of vertical displacement, and
height controls the amount of horizontal displacement.
The defaults values of width and height will introduce noise in both directions.

# Exercise 3.8.3

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```

The geom geom_jitter() adds random variation to the locations points of the graph. In other words, it “jitters” the locations of points slightly. This method reduces overplotting since two points with the same location are unlikely to have the same random variation.

The geom geom_count() sizes the points relative to the number of observations. Combinations of (x, y) values with more observations will be larger than those with fewer observations. The geom_count() geom does not change x and y coordinates of the points. However, if the points are close together and counts are large, the size of some points can itself create overplotting. 

# Exercise 3.8.4

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_boxplot()
```

The default position for geom_boxplot() is "dodge2", which is a shortcut for position_dodge2. This position adjustment does not change the vertical position of a geom but moves the geom horizontally to avoid overlapping other geoms.


# Coordinate systems

```{r}
# Coordinate flip
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
```

```{r}
library(maps)
nz <- map_data("nz")

ggplot(data = nz, mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")

ggplot(data = nz, mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()

```

```{r}
# Coordinate polar
bar <- ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut),
           show.legend = FALSE,
           width = 1) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

# 3.9.1 Exercises

```{r}
ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar()

ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar()

ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
```

# Exercise 3.9.2

The labs function adds axis titles, plot titles, and a caption to the plot.

# Exercise 3.9.3

The coord_map() function uses map projections to project the three-dimensional Earth onto a two-dimensional plane. By default, coord_map() uses the Mercator projection. This projection is applied to all the geoms in the plot. The coord_quickmap() function uses an approximate but faster map projection. This approximation ignores the curvature of Earth and adjusts the map for the latitude/longitude ratio. The coord_quickmap() project is faster than coord_map() both because the projection is computationally easier, and unlike coord_map(), the coordinates of the individual geoms do not need to be transformed.

# Exercise 3.9.4

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed()
```

The function coord_fixed() ensures that the line produced by geom_abline() is at a 45-degree angle. A 45-degree line makes it easy to compare the highway and city mileage to the case in which city and highway MPG were equal.