---
title: "Data transformation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prerequisites

```{r}
library(nycflights13)
library(tidyverse)
```

# Filter rows with filter()

```{r}
jan1 <- filter(flights, month == 1, day == 1)
(dec25 <- filter(flights, month == 12, day == 25))
```

```{r}
filter(flights, month == 11 | month == 12)
nov_dec <- filter(flights, month %in% c(11,12))
```

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <=120, dep_delay<=120)
```

```{r}
df <- tibble(x = c(1,NA,3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```

# Exercise 5.2.1

```{r}
filter(flights, arr_delay >= 120)
filter(flights, dest %in% c("IAH", "HOU"))
filter(flights, carrier %in% c("UA", "AA", "DL"))
filter(flights, month %in% c(7,8,9))
filter(flights, arr_delay >= 120 & dep_delay <= 0)
filter(flights, dep_delay >= 60 & dep_delay - arr_delay > 30)
filter(flights,  dep_time == 2400 | dep_time <= 600)
```

# Exercise 5.2.2

```{r}
filter(flights, between(month, 7, 9))
```

# Exercise 5.2.3

```{r}
filter(flights, is.na(dep_time))
```

Notably, the arrival time (arr_time) is also missing for these rows. These seem to be cancelled flights.

# Exercise 5.2.4

```{r}
NA ^ 0
NA | TRUE
FALSE & NA
NA * 0
```

# Arrange rows with arrange()

```{r}
arrange(flights, year, month , day)
```

```{r}
arrange(flights, desc(dep_delay))
```

```{r}
df <- tibble(x = c(5,2,NA))
arrange(df, x)
arrange(df, desc(x))
```

# Exercise 5.3.1 

```{r}
arrange(flights, desc(is.na(dep_time)))
```

# Exercise 5.3.2

```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```

The most delayed flight was HA 51, JFK to HNL, which was scheduled to leave on January 09, 2013 09:00.
Flight B6 97 (JFK to DEN) scheduled to depart on December 07, 2013 at 21:23 departed 43 minutes early.

# Exercise 5.3.3

```{r}
fastest_flights <- mutate(flights, mph = distance / air_time * 60)
fastest_flights <- select(fastest_flights, mph, distance, air_time, flight, origin, dest, year, month, day)
head(arrange(fastest_flights, desc(mph)))
```

# Exercise 5.3.4

```{r}
arrange(flights, desc(distance))
arrange(flights, distance)
```

# Select columns with select()

```{r}
select(flights, year, month, day)
select(flights, year:day)
select(flights, -(year:day))
```

```{r}
rename(flights, tail_num = tailnum)
```

```{r}
select(flights, time_hour, air_time, everything())
```

# Exercise 5.4.2

```{r}
select(flights, year, month, day, year, year)
```

The select() call ignores the duplication. Any duplicated variables are only included once, in the first location they appear. The select() function does not raise an error or warning or print any message if there are duplicated variables.

# Exercise 5.4.3

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```

The one_of() function selects variables with a character vector rather than unquoted variable name arguments. This function is useful because it is easier to programmatically generate character vectors with variable names than to generate unquoted variable names, which are easier to type.

# Exercise 5.4.4

```{r}
select(flights, contains("TIME"))
```

The default behavior for contains() is to ignore case. This may or may not surprise you. If this behavior does not surprise you, that could be why it is the default. Users searching for variable names probably have a better sense of the letters in the variable than their capitalization. A second, technical, reason is that dplyr works with more than R data frames. It can also work with a variety of databases. Some of these database engines have case insensitive column names, so making functions that match variable names case insensitive by default will make the behavior of select() consistent regardless of whether the table is stored as an R data frame or in a database.


# Add new variables with mutate()

```{r}
flights_sml <- select(flights, year:day, ends_with("delay"), distance, air_time)
mutate(flights_sml, gain = dep_delay - arr_delay, speed = distance / air_time * 60)
```

```{r}
mutate(flights_sml, gain = dep_delay - arr_delay, hours = air_time / 60, gain_per_hours = gain /hours)
```

```{r}
transmute(flights, gain = dep_delay - arr_delay, hours = air_time/60, gain_per_hours = gain / hours)
```

# Useful creation functions

```{r}
transmute(flights, dep_time, hour = dep_time %/% 100, minute = dep_time %% 100)
```

```{r}
x <- 1:10
cumsum(x)
cummean(x)
cummin(x)
cummax(x)
cumprod(x)
```

# Exercise 5.5.1

```{r}
flights_times <- mutate(flights, 
                        dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
                        sched_dep_time_mins = (sched_dep_time %/% 100 * 60 + sched_dep_time %%100) %% 1440)

select(flights_times, dep_time, dep_time_mins, sched_dep_time, sched_dep_time_mins)
```

# Exercise 5.5.2

```{r}
flights_airtime <- mutate(flights, 
                          dep_time = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
                          arr_time = (arr_time %/% 100 * 60 + arr_time %% 100) %% 1440,
                          air_time_diff = air_time + dep_time - arr_time,
                          )

nrow(filter(flights_airtime, air_time_diff != 0))
```

I expect that air_time is the difference between the arrival (arr_time) and departure times (dep_time). In other words, air_time = arr_time - dep_time.
It turns out that there are many flights for which air_time != arr_time - dep_time.

# Exercise 5.5.3

```{r}
flights_deptime <- mutate(flights, 
                          dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
                          sched_dep_time_mins = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440,
                          dep_delay_diff = dep_delay - dep_time_mins + sched_dep_time_mins)
nrow(filter(flights_deptime, dep_delay_diff != 0))
```

I would expect the departure delay (dep_delay) to be equal to the difference between scheduled departure time (sched_dep_time), and actual departure time (dep_time), dep_time - sched_dep_time = dep_delay.

# Exercise 5.5.4

```{r}
flights_delayed <- mutate(flights,
                          dep_delay_min_rank = min_rank(dep_delay),
                          dep_delay_row_number = row_number(dep_delay),
                          dep_delay_dense_rank = dense_rank(dep_delay))

flights_delayed <- filter(flights_delayed, !(dep_delay_min_rank > 10 | dep_delay_row_number > 10 | dep_delay_dense_rank > 10))

flights_delayed <- arrange(flights_delayed, dep_delay_min_rank)

print(select(
  flights_delayed, month, day, carrier, flight, dep_delay,
  dep_delay_min_rank, dep_delay_row_number, dep_delay_dense_rank
),
n = Inf
)
```

Themin_rank() and dense_rank() assign tied values the same rank, but differ in how they assign values to the next rank. For each set of tied values the min_rank() function assigns a rank equal to the number of values less than that tied value plus one. In contrast, the dense_rank() function assigns a rank equal to the number of distinct values less than that tied value plus one. To see the difference between dense_rank() and min_rank() compare the value of rankme$x_min_rank and rankme$x_dense_rank for x = 10.

# Exercise 5.5.5

```{r}
1:3 + 3:10
```

When adding two vectors recycles the shorter vector’s values to get vectors of the same length.

The code also produces a warning that the shorter vector is not a multiple of the longer vector. A warning is provided since often, but not always, this indicates a bug in the code.


# Grouped summaries with summarise()

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))

delay <- filter(delay, count > 20, dest != "HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(mapping = aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

```{r}
# Doing the same previous code but this time using pipe
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(count = n(),
            dist = mean(distance, na.rm = TRUE),
            delay = mean(arr_delay, na.rm = TRUE)) %>%
  filter(count > 20, dest != "HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(mapping = aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
  
```

```{r}
not_cancelled <- flights %>%
  filter(!is.na(arr_delay), !is.na(dep_delay))

not_cancelled %>% 
  group_by(year, month, day) %>%
  summarise(mean = mean(dep_delay))
```

```{r}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(delay = mean(arr_delay))

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)
```

```{r}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE), 
            n = n())

ggplot(data = delays, mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)
```

```{r}
delays %>%
  filter(n > 25) %>%
  ggplot(mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)
```

```{r}
batting <- as_tibble(Lahman::Batting)

batters <- batting %>%
  group_by(playerID) %>%
  summarise(
    ba = sum(H, na.rm = TRUE)/sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>% 
  filter(ab > 100) %>%
  ggplot(mapping = aes(x = ab, y = ba)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
not_cancelled %>% 
  group_by(year, month, day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))
```

```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))
```

```{r}
not_cancelled %>%
  count(tailnum, wt = distance)
```

# Grouping by multiple variables

```{r}
daily <- group_by(flights, year, month, day)
(per_day <- summarise(daily, flights = n()))
(per_month <- summarise(per_day, flights = sum(flights)))
(per_year <- summarise(per_month, flights = sum(flights)))
```

# Ungrouping

```{r}
daily %>% 
  ungroup() %>%
  summarise(flights = n())
```

# 5.6.7 Exercises

In many scenarios, arrival delay is more important. In most cases, being arriving late is more costly to the passenger since it could disrupt the next stages of their travel, such as connecting flights or scheduled meetings.
If a departure is delayed without affecting the arrival time, this delay will not have those affects plans nor does it affect the total time spent traveling. This delay could be beneficial, if less time is spent in the cramped confines of the airplane itself, or a negative, if that delayed time is still spent in the cramped confines of the airplane on the runway.

Variation in arrival time is worse than consistency. If a flight is always 30 minutes late and that delay is known, then it is as if the arrival time is that delayed time. The traveler could easily plan for this. But higher variation in flight times makes it harder to plan.

# Exercise 5.6.2

```{r}
not_cancelled <- flights %>%
  filter(!is.na(arr_delay), !is.na(dep_delay))

not_cancelled %>%
  group_by(dest) %>%
  summarise(destination = n())

not_cancelled %>%
  count(dest)


```

```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(dist = sum(distance))

not_cancelled %>%
  count(tailnum, wt = distance)
```

# Exercise 5.6.3

```{r}
filter(flights, !is.na(dep_delay), is.na(arr_delay)) %>%
  select(dep_time, arr_time, sched_arr_time, dep_delay, arr_delay)
```

If a flight never departs, then it won’t arrive. A flight could also depart and not arrive if it crashes, or if it is redirected and lands in an airport other than its intended destination. So the most important column is arr_delay, which indicates the amount of delay in arrival.

# Exercise 5.6.4

```{r}
cancelled_per_day <- flights %>%
  mutate(cancelled = (is.na(dep_delay)  | is.na(arr_delay))) %>%
  group_by(year, month, day) %>%
  summarise(cancelled_num = sum(cancelled),
            flights_num = n())

ggplot(data = cancelled_per_day) +
  geom_point(mapping = aes(x = flights_num, y = cancelled_num))
```

Plotting flights_num against cancelled_num shows that the number of flights cancelled increases with the total number of flights.

```{r}
cancelled_and_delays <-flights %>%
  mutate(cancelled = (is.na(dep_delay) | is.na(arr_delay))) %>%
  group_by(year, month, day) %>%
  summarise(cancelled_prop = mean(cancelled),
            avg_dep_delay = mean(dep_delay, na.rm = TRUE),
            avg_arr_delay = mean(arr_delay, na.rm = TRUE)
            )

ggplot(data = cancelled_and_delays) +
  geom_point(mapping = aes(x = avg_dep_delay, y = cancelled_prop))

ggplot(data = cancelled_and_delays) +
  geom_point(mapping = aes(x = avg_arr_delay, y = cancelled_prop))
```

There is a strong increasing relationship between both average departure delay and
and average arrival delay and the proportion of cancelled flights.

# Exercise 5.6.6

The sort argument to count() sorts the results in order of n. You could use this anytime you would run count() followed by arrange().


# Grouped mutates (and filters)

```{r}
flights_sml %>%
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```

```{r}
popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n() > 365)
popular_dests
```

# Exercise 5.7.1

```{r}
tibble(
  x = 1 : 9,
  group = rep(c("a","b","c"), each = 3)
) %>%
  mutate(x_mean = mean(x)) %>%
  group_by(group) %>%
  mutate(x_mean2 = mean(x))
```

```{r}
tibble(
  x = 1 : 9,
  group = rep(c("a", "b", "c"), each = 3)
) %>%
  group_by(group) %>%
  mutate(
    lag_x = lag(x),
    lead_x = lead(x)
  )
```

```{r}
tibble(
  x = 1:9,
  group = rep(c("a","b","c"), each = 3)
) %>%
  mutate(x_cumsum = cumsum(x)) %>%
  group_by(group) %>%
  mutate(x_cumsum2 = cumsum(x))
```

```{r}
tibble(
  x = 1:9,
  group = rep(c("a", "b", "c"), each = 3) 
) %>%
  mutate(min_rank = min_rank(x)) %>%
  group_by(group) %>%
  mutate(min_rank2 = min_rank(x))
```

```{r}
tibble(
  x = runif(9),
  group = rep(c("a","b","c"), each = 3)
) %>%
  group_by(group) %>%
  arrange(x) %>%
  mutate(lag = lag(x))
```

# Exercise 5.7.2

```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = (!is.na(arr_time) & arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(on_time) == 1)

flights %>%
  group_by(tailnum) %>%
  summarise(arr_delay = mean(arr_delay), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(desc(arr_delay)) == 1)
```

# Exercise 5.7.3

```{r}
flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)
```

# Exercise 5.7.4

```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    prop_arr_delay = arr_delay / arr_delay_total
  ) %>%
  select(dest, month, day, dep_time, carrier, flight, arr_delay, prop_arr_delay) %>%
  arrange(dest, desc(prop_arr_delay))
```

```{r}
flights %>% 
  filter(arr_delay > 0) %>%
  select(dest, origin, carrier, flight) %>%
  summarise(arr_delay = sum(arr_delay)) %>%
  group_by(dest) %>%
  mutate(arr_delay_prop = arr_delay/sum(arr_delay)) %>%
  arrange(dest, desc(arr_delay_prop)) %>%
  select(carrier, flight, origin, dest, arr_delay_prop)
```

# Exercise 5.7.5

```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay) | !is.na(dep_delay_lag))

lagged_delays
```

```{r}
lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(mapping = aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0,1500, by = 120)) +
  labs(y = "Departure Delay", y = "Previous Departure Delay")  
```

This plots the relationship between the mean delay of a flight for all values of the previous flight. For delays less than two hours, the relationship between the delay of the preceding flight and the current flight is nearly a line. After that the relationship becomes more variable, as long-delayed flights are interspersed with flights leaving on-time. After about 8-hours, a delayed flight is likely to be followed by a flight leaving on time. 

```{r}
lagged_delays %>%
  group_by(origin, dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(mapping = aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  facet_wrap(~ origin, ncol = 1) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```

# Exercise 5.7.6

```{r}

```

