---
title: "Tibbles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prerequisites

```{r}
library(tidyverse)
```


# 10.2 Creating tibbles

```{r}
as_tibble(iris)

tibble(
  x = 1:5,
  y = 1,
  z = x^2 + y
)

tb <- tibble(
  `:)` = "smile",
  ` ` = "space",
  `2000` = "number"
)
tb

tribble(
  ~x, ~y, ~z,
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

# 10.3 Tibbles vs. data.frame

```{r}
tibble(
  a = lubridate:: now() + runif(1e3) *86400,
  b = lubridate:: today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

```{r}
nycflights13::flights %>%
  print(n = 10, width = Inf)
```

# 10.3.2 Subsetting

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df

# Extract by name
df$x
df[["x"]]

# Extract by position
df[[1]]

# Extract using pipe
df %>% .$x
df %>% .[["x"]]
```

# Exercise 10.1

```{r}
mtcars
# When we print mtcars, it prints all the columns.

as_tibble(mtcars)
# But when we first convert mtcars to a tibble using as_tibble(), it prints on the first ten observations. There are also some other differences in formatting of the printed data frame.
```

# Exercise 10.2

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[,"xyz"]
df[,c("abc", "xyz")]

tb <- as_tibble(df)
tb$x
tb[,"xyz"]
tb[,c("abc", "xyz")]
```

The $ operator will match any column name that starts with the name following it. Since there is a column named xyz, the expression df$x will be expanded to df$xyz. This behavior of the $ operator saves a few keystrokes, but it can result in accidentally using a different column than you thought you were using.

With data.frames, with [ the type of object that is returned differs on the number of columns. If it is one column, it won’t return a data.frame, but instead will return a vector. With more than one column, then it will return a data.frame. This is fine if you know what you are passing in, but suppose you did df[ , vars] where vars was a variable. Then what that code does depends on length(vars) and you’d have to write code to account for those situations or risk bugs.

# Exercise 10.3

You can use the double bracket, like df[[var]]. You cannot use the dollar sign, because df$var would look for a column named var.

# Exercise 10.4

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

# 1
annoying[["1"]]
annoying$`1`

# 2
ggplot(annoying, aes(x = `1`, y = `2`)) +
  geom_point()

# 3
mutate(annoying, `3` = `2` / `1`)
annoying[["3"]] <- annoying[["2"]] / annoying[["1"]]
annoying[["3"]] <- annoying$`2` / annoying$`1`

# 4
annoying <- rename(annoying, "one" = `1`, "two" = `2`, "three" = `3`)
glimpse(annoying)
```

# Exercise 10.5

```{r}
enframe(c(a = 1, b = 2, c = 3))
```

The function tibble::enframe() converts named vectors to a data frame with names and values

# Exercise 10.6

The help page for the print() method of tibble objects is discussed in ?print.tbl. The n_extra argument determines the number of extra columns to print information for.
