---
title: "R - Week 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1 : Basic statistics with R

Basic statistics computation with R built-in functions.

```{r}
x = c(4,8,23,2,16,7)

mean(x) # Compute the mean
median(x) # Middle value
length(x) # Number of value in vector or list
range(x) # Largest and smallest value
sd(x) # Standard Deviation
var(x) # Variance

sort(x) # Arange vector or list in right order

summary(x) # Descriptive statistics (range, quartile, mean, median)
```

## Part 2 : Round numbers

Ceiling function (positive values)

```{r}
ceiling(0)
ceiling(0.1)
ceiling(1)
ceiling(1.1)
ceiling(1.5)
```

Ceiling function (negative values)

```{r}
ceiling(-0.1)
ceiling(-1)
ceiling(-1.1)
ceiling(-1.5)
ceiling(-1.9)
ceiling(-2)
```

Floor function (positive values)

```{r}
floor(1)
floor(1.1)
floor(1.5)
floor(1.7)
floor(1.9)
```

Floor function (negative values)

```{r}
floor(-0.1)
floor(-0.5)
floor(-0.9)
floor(-1)
floor(-1.1)
```

Trunc function (positive values)

```{r}
trunc(1)
trunc(1.1)
trunc(1.5)
trunc(1.7)
trunc(1.9)
trunc(2)
```

Trunc function (negative values)

```{r}
trunc(-1)
trunc(-1.1)
trunc(-1.5)
trunc(-1.7)
trunc(-1.9)
trunc(-2)
```

Round function (positive values)

```{r}
round(1)
round(1.1)
round(1.5)
round(1.7)
round(1.9)
```

Round function (negative values)

```{r}
round(-1)
round(-1.1)
round(-1.5)
round(-1.7)
round(-1.9)
```

Round function (with digits)

```{r}
round(1, digits = 0)
round(1.123, digits = 1)
round(-1.534, digits = 2)
round(-1.778, digits = 3)
round(-1.966, digits = 2)
```

Signif function 

```{r}
signif(1.949, digits = 2)
signif(2.453, digits = 1)
signif(-1.357, digits = 0)
signif(3.5, digits = 1)
signif(3.5, digits = 0)
signif(3.500, digits = 2)
```

## Part 3 : Exploring data graphically

```{r}
# Explore data sets available on R
data()
```

```{r}
# Statistical analysis
summary(cars)
```

```{r}
# Graph plotting
plot(cars)
```

```{r}
# Speed ditribution in cars dataset
hist(cars$speed)
```

```{r}
# Distance distribution in cars dataset
hist(cars$dist)
```

```{r}
# Kernal density plots

plot(density(cars$speed)) # Cars speed
plot(density(cars$dist)) # Cars distance
```

# Data visualization with R

```{r}
# Import Thai tourism dataset by reding CSV file
thai_tourist = read.csv("Thaitourism1.csv", header = TRUE)
```

```{r}
# Describe characteristics of the dataset
names(thai_tourist)
head(thai_tourist)
tail(thai_tourist)
str(thai_tourist)
summary(thai_tourist)
```

```{r}
# Import Thai tourism dataset by reding CSV file
thai_tourist_full = read.csv("Thaitourism2.csv", header = TRUE)
```

```{r}
# Describe characteristics of the dataset
names(thai_tourist_full)
head(thai_tourist_full)
tail(thai_tourist_full)
str(thai_tourist_full)
summary(thai_tourist_full)
```

```{r}
# Contain only 2016 data
thai_2016 = thai_tourist[thai_tourist$Year==2016,]
thai_2016
```

```{r}
# Contain only UK nationality
thai_UK = thai_tourist_full[thai_tourist_full$nationality=="UnitedKingdom",]
thai_UK
```

### Plot() function

```{r}
plot(thai_2016$Region, thai_2016$Tourists_1000s)
plot(thai_2016$Region, thai_2016$Tourists_1000s, type = "l", lwd = 1, lty = 3)
plot(thai_tourist$Year, thai_tourist$Tourists_1000s, pch = 20, cex = 2, col = thai_tourist$Region)
```

### Barplot() function

```{r}
barplot(thai_2016$Tourists_1000s, names.arg = thai_2016$Region)
barplot(thai_2016$Tourists_1000s, names.arg = thai_2016$Region, horiz = TRUE)
barplot(thai_2016$Tourists_1000s, names.arg = thai_2016$Region, col = rainbow(8))
```

### Hist() function

```{r}
hist(thai_UK$tourists, labels = TRUE)
hist(thai_UK$tourists, breaks = 16)
hist(thai_UK$tourists, freq = FALSE)
hist(thai_UK$tourists, xlab = "N° ok UK tourists per month",labels = TRUE, col = rainbow(8))
```

### Boxplot() function

```{r}
thai_europe = droplevels(subset(thai_tourist_full, region=="Europe"))
thai_europe
```

```{r}
boxplot(thai_europe$tourists, data = thai_europe)
boxplot(tourists ~ nationality, data = thai_europe, col = rainbow(16))
boxplot(tourists ~ nationality, data = thai_europe, col = rainbow(16), horizontal = TRUE)
boxplot(tourists ~ nationality, data = thai_europe, col = rainbow(16), subset= nationality %in% c("France", "Russia", "Germany"))
```

### Pie() function

```{r}
pie(thai_2016$Tourists_1000s, labels = thai_2016$Region, col = rainbow(8))
```

```{r}
percent = round(100*thai_2016$Tourists_1000s/sum(thai_2016$Tourists_1000s),1)
percent = paste(thai_2016$Region,"-",percent,"%") # add percents to labels
pie(thai_2016$Tourists_1000s, labels = percent, col = rainbow(8))
```

# ggplot2

```{r}
# install ggplot2 packages
install.packages("ggplot2",repos = "http://cran.us.r-project.org")
library(ggplot2)
```

### Scatterplot with qplot() function

```{r}
qplot(Region, Tourists_1000s, data = thai_tourist, color = Year)
qplot(Region, Tourists_1000s, data = thai_tourist, color = Year, size = I(4))
```

### Boxplot with qplot() function

```{r}
qplot(Region, Tourists_1000s, data = thai_tourist, geom = "boxplot", fill = "red")
```

### Barplot with qplot() function

```{r}
qplot(tourists, data = thai_UK, geom = "histogram", fill = "red")
```

### Density with qplot() function

```{r}
qplot(tourists, data = thai_UK, geom = "density", fill = "red")
```

